Transparent Standby for Low-Power, Resource-Constrained Embedded Systems

### 2018-02-16 LCTES/body.tex
Major review

\section{Introduction}
\label{sec.introduction}

According to the International Energy Agency (IEA), the number of
network-connected devices is expected to reach to 50 billion by 2020 with
the expansion of the Internet of Things (IoT).~\cite{iea.data}
%
However, most of the energy to power these devices will be consumed in
\emph{standby mode}, i.e., when they are neither transmitting or processing
data.
%
For instance, standby power currently accounts for 10--15\% of residential
electricity consumption, and $CO_2$ emissions related to standby are equivalent
to those of 1 million cars~\cite{iea.data,standby.australia}.
%
The projected growth of IoT devices, together with the surprising effects of
standby consumption, made network standby efficiency one of the six
pillars of G20's \emph{Energy Efficiency Action Plan}%
\footnote{G20's Energy Efficiency Action Plan: \url{https://www.iea-4e.org/projects/g20}}.

@@

Our approach lies at the bottom of the software development
layers---transparent programming language mechanisms---meaning that \emph{all}
applications take advantage of low-power standby modes automatically, without
extra programming efforts.
%
We extend the synchronous programming language
\CEU~\cite{ceu.sensys13,ceu.tecs17} with interrupt service routines (ISRs) and
a simple power management runtime (PMR).
%
Each supported microcontroller requires bindings in C for the ISRs and PMR, and
each peripheral requires a driver in \CEU.
This is a one-time process and is typically packaged and distributed in a SDK.
%
All new applications built on top of these drivers will take advantage of
standby automatically.
%
As a proof of concept, we support the 8-bit \emph{AVR/ATmega} and 32-bit
\emph{ARM/Cortex-M0} microcontrollers, and a variety of peripherals (e.g., ADC,
SPI, USART, nRF24L01 transceiver).
The project is open-source.%
\footnote{https://github.com/fsantanna/ceu-arduino/}

 Thank you, I will improve the page.

@@

In comparison to promises/futures, it provides lexical parallel constructs, allowing the branches to share local variables and, more importantly, supporting safe abortion of code (with the "par/or").

- application on IoT
- gains??

- structured programming
- lexical scope
- automatic power management: standby as well as enabled/disabled, powered, switched

needs a requires a energy-aware driver

process for ISR bindings and
A infra-structure programmer writes device drivers for
Each peripheral requires a driver
An architecture
Infra
Once a driver is written, all applications

- synchronous languages / \CEU

- except drivers, which are also written in Ceu

I have been working in the design of \CEU, a new reactive programming language
targeting resource-constrained embedded systems, for the past 8
years~\cite{ceu.sensys11,ceu.tr,ceu.sensys13,ceu.rem13,ceu.phd,ceu.rebls14,
ceu.mod15,ceu.rebls15,ceu.terra,ceu.media.webmedia16,ceu.tecs17}.
%
\CEU is grounded on the synchronous concurrency model, which has been
successfully adopted in the context of hard real-time systems such as avionics
and automobiles industry since the 80's~\cite{rp.twelve}.
%
The synchronous model trades power for reliability and has a simpler model
of time that suits most requirements of IoT applications.
%
On the one hand, this model cannot directly express time-consuming
computations, such as compression and cryptography algorithms, which are
typically either absent or delegated to auxiliary chips in the context of the
IoT.
%
On the other hand, all reactions to the external world are guaranteed to be
computed in bounded
time, ensuring that applications always reach an idle state amenable to standby
mode.

- results
    - atmega 328p/2560
    - arm cortex-m0
    - how much efficiency?
%
We expect that existing energy-unaware applications will benefit from savings
in the order of 50\% based on IEA's estimates considering current standby
technologies~\cite{iea.data}, and previous third-party work on transparent energy
awareness~\cite{wsn.tos.2}.

- limitations

- paper structure

In Section~\ref{sec.ceu}, we introduce the synchronous model together with the
programming language \CEU.
In Section~\ref{sec.ext}, we present the language extensions that support
transparent standby.
In Section~\ref{sec.eval}, we evaluate .

\section{The Synchronous Language CEU}
\label{sec.ceu}

@@

>>>>

\label{sec.introduction}

According to the International Energy Agency (IEA), the number of
network-connected devices is expected to reach 50 billion by 2020 with the
expansion of the Internet of Things (IoT)~\cite{iea.data}.
%
However, most of the energy to power these devices will be consumed in
\emph{standby mode}, i.e., when they are neither transmitting or processing
data.
%
For instance, standby power currently accounts for 10--15\% of residential
electricity consumption, and $CO_2$ emissions related to standby are equivalent
to those of 1 million cars~\cite{iea.data,standby.australia}.
%
The projected growth of IoT devices, together with the surprising effects of
standby consumption, made network standby efficiency one of the six
pillars of the G20's \emph{Energy Efficiency Action Plan}%
\footnote{G20's Energy Efficiency Action Plan: \url{https://www.iea-4e.org/projects/g20}}.

@@

Our approach lies at the bottom of the software development
layers---programming language mechanisms---meaning that \emph{all} applications
take advantage of low-power standby modes automatically, without extra
programming efforts.
%
We extend the synchronous programming language
\CEU~\cite{ceu.sensys13,ceu.tecs17} with support for interrupt service routines
(ISRs) and with a simple power management runtime (PMR).
%
Each supported microcontroller requires bindings in C for the ISRs and PMR, and
each peripheral requires a driver in \CEU.
These are a one-time procedures and are typically packaged and distributed in a
software development kit (SDK).
%
Then, all new applications built on top of these drivers take advantage of
standby automatically.
%
As a proof of concept, we provide an open source SDK %
%\footnote{https://github.com/fsantanna/ceu-arduino/}
with support for 8-bit
\emph{AVR/ATmega} and 32-bit \emph{ARM/Cortex-M0} microcontrollers, and a
variety of peripherals, such as GPIO, ADC, USART, SPI, and the nRF24L01
transceiver.

We developed a number of applications using these peripherals concurrently and
could verify that the applications remain in deepest standby modes for longest
periods of time.
%
We also compare the structure of programs in \CEU and
Arduino~\cite{arduino.book}, whose primary goal is to reduce the barrier of
adoption for non-technical users (e.g., designers and artists).
We show that we can keep the intended sequential reasoning of Arduino even when
applications require non-trivial concurrent behaviors.

% - standby as well as enabled/disabled, powered, switched
% - application on IoT
% - results
%   - atmega 328p/2560
%   - arm cortex-m0
%   - how much efficiency?
% - limitations

%In Section~\ref{sec.ceu}, we introduce the structured synchronous model and the
%programming language \CEU.
%In Section~\ref{sec.apps}, we evaluate .
%In Section~\ref{sec.ext}, we present the language extensions that support
%transparent standby.

\section{The Structured Synchronous Programming Language \CEU}

\begin{comment}
- structured programming
- lexical scope

@@

In comparison to promises/futures, it provides lexical parallel constructs, allowing the branches to share local variables and, more importantly, supporting safe abortion of code (with the "par/or").
\end{comment}

\CEU is a Esterel-based\cite{ceu.tecs17} reactive programming language
targeting resource-constrained embedded systems~\cite{ceu.sensys13}.
%
It is grounded on the synchronous concurrency model, which has been
successfully adopted in the context of hard real-time systems such as avionics
and automobiles industry since the 80's~\cite{rp.twelve}.
%
The synchronous model trades power for reliability and has a simpler model
of time that suits most requirements of IoT applications.
%
On the one hand, this model cannot directly express time-consuming
computations, such as compression and cryptography algorithms, which are
typically either absent or delegated to auxiliary chips in the context of the
IoT.
%
On the other hand, all reactions to the external world are guaranteed to be
computed in bounded time, ensuring that applications always reach an idle state
amenable to standby mode.
%
Overall, \CEU aims to offer a concurrent, safe, and expressive alternative to C
with the characteristics that follow:
%
\begin{description}
\item [Reactive:] code only executes in reactions to events.
\item [Structured:] programs use structured control mechanisms, such as
    \code{await} (to suspend a line of execution), and \code{par} (to combine
    multiple lines of execution).
\item [Synchronous:] reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{description}
%
Structured reactive programming lets developers write code in direct style,
recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.

\subsection*{A Motivating Example}
\label{sec.ceu.example}

{\linespread{1}
\begin{figure}[t]
\begin{minipage}[t]{0.49\linewidth}
\begin{lstlisting}[xrightmargin=0.5cm]
while (1) {
  delay(1000);
  int v =
    analogRead(A0);
  radioWrite(v);
}
\end{lstlisting}
\centering\small{[a] Version in Arduino}
\end{minipage}
%
\begin{minipage}[t]{0.49\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=-0.2cm]
loop do
  await 1s;
  var int v =
    await AnalogRead(A0);
  await RadioWrite(v);
end
\end{lstlisting}
\centering\small{[b] Version in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Sequence of I/O operations running in a loop.
\label{lst.direct}
}
\end{figure}
}

Figure~\ref{lst.direct}.a shows a simple, easy-to-read program chunk in Arduino
that executes forever in a loop a sequence of operations as follows:
    waits for 1 second (ln. 2),
    performs an analog to digital conversion (ln. 3--4), and
    broadcasts the value read (ln. 5).
%
Figure~\ref{lst.direct}.b shows the same chunk in \CEU, with a noteworthy
difference that operations that interact with the environment and take time use
the \code{await} keyword. % and can be easily identified in the program.
%
The traditional structured paradigm encouraged in Arduino (with blocks, loops,
and sequences) allows for simple and readable code, avoiding the complexity of
dealing with ISRs.
%
However, the use of blocking operations, such as \code{delay(1000)}, prevents
that other operations execute concurrently.

Suppose we now want that, at any time, receiving a message via radio should
immediately abort the loop in Figure~\ref{lst.direct}.a.
%
Since the message might arrive concurrently with any of the blocking
operations, we need to change the structure of the program.
%
Figure~\ref{lst.inversion}.a changes the blocking operation \code{delay} to the
polling operation \code{millis}, which immediately returns the number of
milliseconds since the reset.
Now, we start by registering the current time (ln. 1--2) and, on each loop
iteration, we recheck the time to see if one second has elapsed (ln. 7--9).
Since these operations are non-blocking, we can intercalate the execution with
checks for message arrivals (ln. 4--6).
If the time is up, we start counting it again (ln. 10) before proceeding to the
original operations in sequence (ln. 11--13).
%
The original structured style has been drastically violated to accommodate
concurrency.
In the example, we only adapted the \code{delay} operation, but the other
blocking operations (\code{analogRead} and \code{radioWrite}) would also need
to be changed to achieve maximum concurrency.
%
Alternatively, we could resort to ISRs or implement an event-driven
scheduler to handle the operations~\cite{tos.nesc}, but ultimately, the
program readability would still be compromised.

The program in Figure~\ref{lst.inversion}.b in \CEU extends the one in
Figure~\ref{lst.direct}.b to accommodate concurrency.
%
The original code remains unmodified (ln. 4--9) and concurrency is achieved
through the \code{par/or} construct, which creates two lines of execution and
terminates when either of them terminates, aborting the other automatically.
%
This approach preserves the sequential, easy-to-read style while accommodating
concurrency seamlessly.

\begin{figure}[t]
\begin{minipage}[t]{0.49\linewidth}
\begin{lstlisting}[xrightmargin=0.5cm]
uint32_t prv =
  millis();
while (1) {
  if (radioAvail()) {
    break;
  }
  uint32_t cur =
    millis();
  if (cur>prv+1000) {
    prv = cur;
    int v =
      analogRead(A0);
    radioWrite(v);
  }
}
\end{lstlisting}
\centering\small{[a] Version in Arduino}
\end{minipage}
%
\begin{minipage}[t]{0.49\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=-0.2cm]
par/or do
  await RadioAvail();
with
  loop do
    await 1s;
    var int v =
      await AnalogRead(A0);
    await RadioWrite(v);
  end
end

.
\end{lstlisting}
\centering\small{[b] Version in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Achieving concurrency between I/O operations.
\label{lst.inversion}
}
\end{figure}

\subsection*{Standby Considerations}
\label{sec.ceu.standby}

Another advantage of the example in Figure~\ref{lst.inversion}.b is that the
structure of the program indicates which peripherals are active at a given
time.
%
For instance, when the program is awaiting concurrently in lines 2 and 7,
only the radio and ADC can awake the program.
Hence, the language runtime can choose the most energy-efficient sleep mode
that allows these peripherals to awake the microcontroller from appropriate
interrupts.
%
Since the semantics of \CEU forces the program to always reach \code{await}
statements in all active lines of execution, it is always possible to put the
microcontroller into the optimal sleep mode.

This way, when an application reaches an idle state, the language has precise
information about which events can awake that application.

By design, all lines of execution in \CEU always reach an \code{await} at the
end of a reaction to an event (otherwise, the application does not
compile~\cite{ceu.sensys13}).
%
This not only allows the application to enter standby mode, but effectively use
the deepest sleeping level considering all possible awaking events.
While in standby, only a hardware interrupt associated with the events can
awake the application, making it sleep for longest possible periods of time.

@@

\begin{figure}[t]
\begin{lstlisting}[numbers=left]
output (int pin) ANALOG_REQUEST do
    <...>   // port manipulation to start the conversion
end

spawn async/isr [CEU_ADC_IRQ_N, 0] do
    var int value = <...>; // port with the value read
    emit ANALOG_DONE(value);
end

code AnalogRead (var int pin) -> int do
    {ceu_pm_set(CEU_PM_ANALOG, 1);}
    do finalize with
        {ceu_pm_set(CEU_PM_ANALOG, 0);}
    end

    emit ANALOG_REQUEST(pin);
    var int value = await ANALOG_DONE;

    escape value;
end
\end{lstlisting}
%\rule{8.4cm}{0.37pt}
\caption{ \CEU driver for the analog-to-digital converter.
\label{lst.adc}
}
\end{figure}

### 2018-02-16 LCTES/lctes.tex
Major review

\usepackage{booktabs} % For formal tables

@@

\begin{abstract}
This paper provides a sample of a \LaTeX\ document which conforms,
somewhat loosely, to the formatting guidelines for
ACM SIG Proceedings.
\end{abstract}

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{serra,my,standby}

>>>>

\usepackage{booktabs} % For formal tables
\usepackage{xspace}

\usepackage{listings}
\lstdefinelanguage{ceu}{%
  language=C,
  morekeywords={%
    and, uint32_t, code,
    await, bool, call, class, data, define, deterministic, do, each, else,
    emit, end, escape, event, every, finalize, hor, implementation, in,
    input, interface, loop, min, native, new, nohold, not, or, output, par,
    pool, pure, return, signal, spawn, tag, then, this, traverse, until,
    var, watching, when, with},
}
\lstset{
  basicstyle=\ttfamily,
  captionpos=b,
  columns=flexible,
  commentstyle=\rmfamily\itshape,
  escapeinside={||},
  frame=tb,
  keepspaces=true,
  keywordstyle=\bfseries,
  language=ceu,
  mathescape=true,
  numbersep=4pt,
  numberstyle=\scriptsize,
  upquote=true,
}

\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\code}[1] {{\small{\texttt{#1}}}}

@@

\begin{abstract}
Standby efficiency for connected devices is one of the priorities of the G20's
\emph{Energy Efficiency Action Plan}.
%
We propose transparent programming language mechanisms to enforce that
applications remain in deepest standby modes for longest periods of time.
%
We extend the synchronous programming language \CEU with support for interrupt
service routines and with a simple power management runtime.
%
We developed device drivers based on these primitives on top of which
applications can be built to take advantage of standby automatically.
%
We also show that programs in \CEU can keep a sequential structure to lower the
barrier of adoption, even when applications require non-trivial concurrent
behaviors.
\end{abstract}

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{my,standby,other}
